Eliminating object prior-bias from sparse-projection tomographic reconstructions

### 2021-09-15 paper/post_tci.tex
Changes made based on comments from Prof.Ajit

\author{Preeti Gopal,
  Sharat Chandran,
  Imants Svalbe,
        and~Ajit Rajwade
\thanks{Preeti Gopal is currently with the Applied Artificial Intelligence Institute, Deakin University}
\thanks{Sharat Chandran and Ajit Rajwade are with the Department of Computer Science and Engineering, IIT Bombay}
\thanks{Imants Svalbe is with the School of Physics and Astronomy, Monash University}}

@@

In conventional data acquisition techniques, tomographic measurements
$\boldsymbol{y}$ are acquired by sampling the physical object
$\boldsymbol{x}$ uniformly with a substantial number of views, ideally
above the Nyquist rate. In such a case when there are sufficient
measurements, reconstruction using conventional filtered
backprojection (FBP) suffices, as seen in Fig.~\ref{fig:story} (first
column, 800 views). The figure shows the ground truth image
(260$\times$260) of naturally growing sprouts at the top left (the
details of this dataset is postponed to Section~\ref{sec:datasets}).

However, reconstruction from reduced views has been made possible by
assuming the data to exhibit certain properties such as smoothness in
image space, sparsity in gradient space~\cite{TV} or sparsity under
certain mathematical transforms~\cite{Donoho,introCS} such as the
Discrete Cosine Transform (DCT).  There are multiple such
regularization priors such as the Compressed
Sensing based l1-ls prior~\cite{my_dicta_paper}. For the purposes of
demonstrating the efficacy of our method, we use the well known total Variation (TV)
regularization prior.
If $\boldsymbol{\mathcal{R}}$ represents the acquisition
model, then the TV-regularized solution is described by one that
minimizes a combination of least squares error and the $TV$-norm of
the object $\boldsymbol{x}$. Mathematically, we choose to minimize
$J_{TV}(\boldsymbol{x})$ where

@@

  $\nabla$ representing the gradient operator.

@@

The methods that use object-specific knowledge are~\cite{Van2015}
and~\cite{Marjolein2016}.  In~\cite{Van2015}, knowledge of attenuation
coefficient of the fluid is used as a prior for reconstructing its
flow through gravel. However, this relies on additional expert domain
knowledge which may not be available or which may be infeasible to
acquire. Other approaches variously use robust principal components
analysis assuming that the object consists of a large static part
(background) and a sparse moving part (foreground)~\cite{HaoGao},
spline-based models for tracking the regions of change in an SIRT
framework~\cite{Van2014}.

@@

\section{Contributions}
\label{sec:contributions}
This paper focuses on few-views reconstruction with an emphasis on
longitudinal studies. In contrast to the object-prior based studies
mentioned above, we reconstruct the current test object without any
assumption of continuity of changes or knowledge of the attenuation
coefficients of the structures. We also do not make any temporal
assumption in terms of time intervals -- prior scans could be months
apart. We use the current measurements from few-views and previous
scans of the same object.

@@

%-------------------------------Dataset----------------------------
\section{Datasets}
\label{sec:datasets}
In this section, we present details of the biological and medical
datasets created and used for this work. For the first two datasets
namely, Potato and Okra, the raw measurements and  corresponding
paramteres for cone-beam imaging model are available to us.
Hence all reconstructions on these datasets were performed directly
from real projection measurements. For the other two datasets namely, Sprouts and
Liver, the raw measurements and the parameters of imaging
model were not available to us (due to propriety concerns) and hence
only reconstructed volumes of real data were at hand; we therefore
generated measurements from these datasets assuming a circular
cone-beam imaging model for 3D and a parallel-beam imaging model for
2D.

@@

 \textbf{Liver:} %We have used this dataset
 %to illustrate an application
%where \textit{both} uniform and spatially-varying priors are
%useful.
TMC is the national comprehensive centre for the prevention,
treatment, education and research in cancer, and is recognized as one
of the leading cancer centres in India. The dataset
(Fig.~\ref{fig:RFA2_test_object-prior}) from this longitudinal medical
study consists of 7 scans taken at different stages of radio-frequency
ablation study of a liver. In such a procedure~\cite{Dong2015}, the
physician inserts a thin needle-like probe into the organ. Repeated CT
scans of the patient are acquired in order to track the movement of
the needle and to ensure that it is reaching the appropriate target
tumor. Once the needle hits the tumor, a high-frequency electric
current is passed through the tip of the probe and this burns the
malignant tumor (ablation).

@@

\textit{Remark:} Liver is interesting because the data comes from live
data from a clinical setting.  The oncologist has to be aware of the
precise position of the needle contrasted with the site of the tumor,
yet we do not want to create new mutant DNA by further bombarding the
patient with X-rays.

@@

$\lambda_1$, $\lambda_2$ are tunable weights given to the
TV-prior and object-prior terms respectively. The unknowns
$\boldsymbol{\x}$ and $\boldsymbol{\alpha}$ are solved by alternately
minimizing $J_{\boldsymbol{\alpha}}(\boldsymbol{\x})$ using a fixed
$\boldsymbol{\alpha}$, and $J_{\boldsymbol{x}}(\boldsymbol{\alpha})$
using the resultant $\boldsymbol{\x}$.
$J_{\boldsymbol{\alpha}}(\boldsymbol{\x})$ is solved for using an
optimal first-order method for large-scale TV regularization presented
in~\cite{TVReg} and available in~\cite{TVReg-lib}.

@@

\begin{equation}
J_{\boldsymbol{x}}(\boldsymbol{\alpha}) \triangleq \lVert\boldsymbol{\boldsymbol{W}(\boldsymbol{x}} - (\boldsymbol{\mu + V\alpha}))\rVert_2^2.
\end{equation}

@@

  \item From $\boldsymbol{y_{Q_i}}$, perform reconstructions of the
    template $Q_i$ using the $M$ different algorithms, for each of the
    $L$ previously scanned objects. Let this set be denoted by $Y
    \triangleq \{\{Y_{i}^j\}_{j=1}^M\}_{i=1}^L$ where $Y^{1}_i =
    Y^{\text{fdk}}_i$, $\forall i \in \{1,..,,L\}$.

@@

The test volume was reconstructed from a partial set of 45 real
cone-beam projections, i.e., $5\%$ of the $900$ projection views from
which ground truth was reconstructed (see remarks in
Sec.~\ref{sec:potatoAndOkra}). 2D reconstructions of one of the slices
is shown in Fig.~\ref{fig:potato_2D_results}. A complete 3D
reconstruction can be found in the supplementary
material~\cite{supp_paper}. The red RoI spans 7 consecutive slices
where the test is different from all of the previously scanned
objects. %The zoomed-in images around the major region of change (red
%RoI) is shown in Fig.~\ref{fig:potato_zoomed_2D_results}. 
\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{blue}{blue}{\includegraphics[width=0.22\columnwidth]{../images/potato/post_tci/testIm_paper.png}}}
\subcaptionbox{FDK, no prior}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/fbp.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/tv_lambdaTV_0.05.png}}
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/weighted_prior_kk_30_lambda_prior_1.100000.png}}
\caption{Reconstruction of potato from 6 projection views--(b) has strong streak artefacts with unclear structure of the potato, (c) largely blurred, and (c) is sharper with significantly reduced streak artefacts.}
\label{fig:potato_2D_results}
\end{figure}

@@

%--------------------------Okra-- results
\subsubsection{\textbf{Okra}}
\label{Sec:okra_spatially_varying}
The test volume was reconstructed from a partial set of 45 real
cone-beam projections, i.e., $10\%$ of the $450$ projection views from which
ground truth was reconstructed (refer Sec.~\ref{sec:datasets}). 2D reconstructions of one of the
slices is shown in Fig.~\ref{fig:okra_2D_results}. A complete 3D
reconstruction can be found in in the supplementary material
~\cite{supp_paper}. The red RoI spans 7 consecutive slices where the
test is different from all of the previously scanned objects. %A
%zoomed-in region around the major region of change (red RoI) is shown
%in Fig.~\ref{fig:okra_zoomed_2D_results}.
As in the test, the
reconstruction by spatially-varying method shows the absence of the
deformity and better removal of sub-sampling artefacts when compared
to FDK and TV. This is also seen in the SSIM values in
Table~\ref{table:okra_ssim}.

\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{cyan}{cyan}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/input_roi.png}}}
\subcaptionbox{FBP, no prior}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/FBP_cropped.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/TV_cropped.png}}
%\subcaptionbox{\mbox{Uniform}\\prior}{\includegraphics[width=0.19\columnwidth]{../images/okra/pca_cropped.png}}\hfill
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/weighted_prior_cropped.png}}
\caption{Reconstruction of okra from 48 projection views
   (b) has streaky artefacts, (c) has blurred structures and (d) is sharper with significantly less streaky artefacts.}
\label{fig:okra_2D_results}
\end{figure}

@@

The test volume was reconstructed from partial set of 45 real
cone-beam projections, i.e., $2.5\%$ of the $1800$ projection views
from which ground truth was reconstructed (refer
Sec.~\ref{sec:datasets}). The selected 3D ground truth of template
volumes, test volume, as well as the 3D reconstructions are shown in
the supplementary material~\cite{supp_paper}. For the sake of
exposition, the red region of interest (RoI) has been culled out from
7 consecutive slices in the 3D volume to indicate new structures;
other changes can be viewed in the video.  2D reconstruction of one of
the slices is shown in Fig.~\ref{fig:sprouts_2D_results}.
Table~\ref{table:sprouts_ssim} shows the improvement in SSIM of the
reconstructed new regions as compared to other methods.
%----------------------------------------------------

\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{yellow}{yellow}{\includegraphics[width=0.22\columnwidth]{../images/sprouts/post_TCI/2D/20_views/testIm_roi.png}}}
\subcaptionbox{FDK, no prior}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/fbp.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/tv_lambdaTV_0.100.png}}
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/weighted_prior_kk_30_lambda_prior_2.100000.png}}
\caption{Reconstruction of sprouts from 20 projection views
   (b) has streaky artefacts, (c) has blurred structures and (d) is sharper with significantly less streaky artefacts.}
\label{fig:sprouts_2D_results}
\end{figure}

>>>>

\author{Preeti Gopal,
  Sharat Chandran,
  Imants Svalbe,
        and~Ajit Rajwade
\thanks{Preeti Gopal is currently with the Applied Artificial Intelligence Institute, Deakin University; p.gopal@deakin.edu.au}
\thanks{Sharat Chandran and Ajit Rajwade are with the Department of Computer Science and Engineering, IIT Bombay; \{sharat, ajitvr\}@cse.iitb.ac.in}
\thanks{Imants Svalbe is with the School of Physics and Astronomy, Monash University; imants.svalbe@monash.edu}}

@@

In conventional data acquisition techniques, tomographic measurements
$\boldsymbol{y}$ are acquired by sampling the physical object
$\boldsymbol{x}$ uniformly with a substantial number of views, ideally
above the Nyquist rate. In such a case when there are sufficient
measurements, reconstruction using conventional filtered
backprojection (FBP) suffices, as seen in Fig.~\ref{fig:story} (first
column, 800 views). The figure shows the ground truth image
(260$\times$260) of naturally growing sprouts at the top left (the
details of this dataset are postponed to Section~\ref{sec:datasets}).

However, reconstruction from reduced views has been made possible by
assuming the data to exhibit certain properties such as smoothness in
image space~\cite{Essam2015}, sparsity in gradient space~\cite{TV} or sparsity under
certain mathematical transforms~\cite{Donoho,introCS} such as the
Discrete Cosine Transform (DCT).  There are multiple such
regularization priors such as the Compressed
Sensing based l1-ls prior~\cite{my_dicta_paper}. For the purposes of
demonstrating the efficacy of our method, we use the well known total Variation (TV)
regularization prior.
If $\boldsymbol{\mathcal{R}}$ represents the acquisition
model, then the TV-regularized solution is described by one that
minimizes a combination of least squares error and the $TV$-norm of
the object $\boldsymbol{x}$. Mathematically, we choose to minimize
$J_{TV}(\boldsymbol{x})$ where

@@

  where $\nabla$ represents the gradient operator.

@@

The methods that use object-specific knowledge are~\cite{Van2015}
and~\cite{Marjolein2016}.  In~\cite{Van2015}, knowledge of attenuation
coefficient of the fluid is used as a prior for reconstructing its
flow through gravel. However, this relies on additional expert domain
knowledge which may not be available or which may be infeasible to
acquire. Other approaches variously use robust principal components
analysis assuming that the object consists of a large static part
(background) and a sparse moving part (foreground)~\cite{HaoGao} or
spline-based models for tracking the regions of change in an SIRT
framework~\cite{Van2014}.

@@

\section{Contributions}
\label{sec:contributions}
This paper focuses on few-views reconstruction with an emphasis on
longitudinal studies. In contrast to the object-prior based studies
mentioned above, we reconstruct the current test object without any
assumption of continuity of changes or knowledge of the attenuation
coefficients of the structures. We also do not make any temporal
assumption in terms of time intervals -- prior scans could be months
apart which is typical in cancer scanning, or could be a few
minutes/seconds apart which is typical in CT-guided surgeries. We use
the current measurements from few-views and previous scans of the same
object.

@@

%-------------------------------Dataset----------------------------
\section{Datasets}
\label{sec:datasets}
In this section, we present details of the biological and medical
datasets created and used for this work. For the first two datasets
namely, Potato and Okra, the raw measurements and  corresponding
parameters for cone-beam imaging model are available to us.
Hence all reconstructions on these datasets were performed directly
from real projection measurements. For the other two datasets namely, Sprouts and
Liver, the raw measurements and the parameters of imaging
model were not available to us (due to propriety concerns) and hence
only reconstructed volumes of real data were at hand; we therefore
generated measurements from these datasets assuming a circular
cone-beam imaging model for 3D and a parallel-beam imaging model for
2D.

@@

 \textbf{Liver:} %We have used this dataset
 %to illustrate an application
%where \textit{both} uniform and spatially-varying priors are
%useful.
TMC is the national comprehensive centre for the prevention,
treatment, education and research in cancer, and is recognized as one
of the leading cancer centres in India. The dataset
(Fig.~\ref{fig:RFA2_test_object-prior}) from this longitudinal medical
study consists of 7 scans taken at different stages of radio-frequency
ablation (RFA) study of a liver. In such a procedure~\cite{Dong2015}, the
physician inserts a thin needle-like probe into the organ. Repeated CT
scans of the patient are acquired in order to track the movement of
the needle and to ensure that it is reaching the appropriate target
tumor. Once the needle hits the tumor, a high-frequency electric
current is passed through the tip of the probe and this burns the
malignant tumor (ablation).  A fairly high amount of radiation is administered to a patient in this RFA procedure~\cite{radiation_dose_RFA}, and hence reducing radiation per scan is highly beneficial here.

@@

\textit{Remark:} This dataset is interesting because it comes from a live
clinical setting.  The oncologist has to be aware of the
precise position of the needle contrasted with the site of the tumor,
yet we do not want to create new mutant DNA by further bombarding the
patient with X-rays~\cite{damage_repeated_CT}.

@@

$\lambda_1$, $\lambda_2$ are tunable weights given to the
TV-prior and object-prior terms respectively. The unknowns
$\boldsymbol{x}$ and $\boldsymbol{\alpha}$ are solved by alternately
minimizing $J_{\boldsymbol{\alpha}}(\boldsymbol{x})$ using a fixed
$\boldsymbol{\alpha}$, and $J_{\boldsymbol{x}}(\boldsymbol{\alpha})$
using the resultant $\boldsymbol{x}$.
$J_{\boldsymbol{\alpha}}(\boldsymbol{x})$ is solved for using an
optimal first-order method for large-scale TV regularization presented
in~\cite{TVReg} and available in~\cite{TVReg-lib}.

@@

\begin{equation}
J_{\boldsymbol{x}}(\boldsymbol{\alpha}) \triangleq \lVert{\boldsymbol{W}(\boldsymbol{x}} - (\boldsymbol{\mu + V\alpha}))\rVert_2^2.
\end{equation}

@@

  \item From $\boldsymbol{y_{Q_i}}$, perform reconstructions of the
    template $Q_i$ using the $M$ different algorithms, for each of the
    $L$ previously scanned objects. Let this set be denoted by $Y
    \triangleq \{\{Y_{i}^j\}_{j=1}^M\}_{i=1}^L$ where $Y^{1}_i =
    Y^{\text{fdk}}_i$, $\forall i \in \{1,..,,L\}$. If the set of projection views for every test is fixed a priori, this set of low quality reconstructions can be produced offline and stored, in order to save computational costs. If new sets of projection views for every test are to be allowed, the low quality reconstructions can still be performed efficiently using parallelization.

@@

The test volume was reconstructed from a partial set of 18 real
cone-beam projections, i.e., $2\%$ of the $900$ projection views from
which ground truth was reconstructed (see remarks in
Sec.~\ref{sec:potatoAndOkra}). 2D reconstructions of one of the slices
is shown in Fig.~\ref{fig:potato_2D_results}. A complete 3D
reconstruction can be found in the supplementary
material~\cite{supp_paper}. The red RoI spans 7 consecutive slices
where the test is different from all of the previously scanned
objects. %The zoomed-in images around the major region of change (red
%RoI) is shown in Fig.~\ref{fig:potato_zoomed_2D_results}. 
\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{blue}{blue}{\includegraphics[width=0.22\columnwidth]{../images/potato/post_tci/testIm_paper.png}}}
\subcaptionbox{FDK, no prior}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/fbp.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/tv_lambdaTV_0.05.png}}
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/potato/post_tci/6_views/weighted_prior_kk_30_lambda_prior_1.100000.png}}
\caption{Reconstruction of potato from 6 projection views--(b) has strong streak artefacts with unclear structure of the potato, (c) largely blurred, and (c) is sharper with significantly reduced streak artefacts. 3D reconstructions from 18 cone-beam views can be seen in the supplementary material~\cite{supp_paper}. }
\label{fig:potato_2D_results}
\end{figure}

@@

%--------------------------Okra-- results
\subsubsection{\textbf{Okra}}
\label{Sec:okra_spatially_varying}
The test volume was reconstructed from a partial set of 30 real
cone-beam projections, i.e., $6.7\%$ of the $450$ projection views from which
ground truth was reconstructed (refer Sec.~\ref{sec:datasets}). 2D reconstructions of one of the
slices is shown in Fig.~\ref{fig:okra_2D_results}. A complete 3D
reconstruction can be found in in the supplementary material
~\cite{supp_paper}. The red RoI spans 7 consecutive slices where the
test is different from all of the previously scanned objects. %A
%zoomed-in region around the major region of change (red RoI) is shown
%in Fig.~\ref{fig:okra_zoomed_2D_results}.
As in the test, the
reconstruction by spatially-varying method shows the absence of the
deformity and better removal of sub-sampling artefacts when compared
to FDK and TV. This is also seen in the SSIM values in
Table~\ref{table:okra_ssim}.

\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{cyan}{cyan}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/input_roi.png}}}
\subcaptionbox{FBP, no prior}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/FBP_cropped.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/TV_cropped.png}}
%\subcaptionbox{\mbox{Uniform}\\prior}{\includegraphics[width=0.19\columnwidth]{../images/okra/pca_cropped.png}}\hfill
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/okra/post_TCI/2D/48_views/weighted_prior_cropped.png}}
\caption{Reconstruction of okra from 48 projection views
   (b) has streaky artefacts, (c) has blurred structures and (d) is sharper with significantly less streaky artefacts. 3D reconstructions from 30 cone-beam views can be seen in the supplementary material~\cite{supp_paper}.}
\label{fig:okra_2D_results}
\end{figure}

@@

The test volume was reconstructed from partial set of 30 
cone-beam projections, i.e., $1.7\%$ of the $1800$ projection views
from which ground truth was reconstructed (refer
Sec.~\ref{sec:datasets}). The selected 3D ground truth of template
volumes, test volume, as well as the 3D reconstructions are shown in
the supplementary material~\cite{supp_paper}. For the sake of
exposition, the red region of interest (RoI) has been culled out from
7 consecutive slices in the 3D volume to indicate new structures;
other changes can be viewed in the video.  2D reconstruction of one of
the slices is shown in Fig.~\ref{fig:sprouts_2D_results}.
Table~\ref{table:sprouts_ssim} shows the improvement in SSIM of the
reconstructed new regions as compared to other methods.
%----------------------------------------------------

\begin{figure}[!h]
\centering
\subcaptionbox{Test}{\fcolorbox{yellow}{yellow}{\includegraphics[width=0.22\columnwidth]{../images/sprouts/post_TCI/2D/20_views/testIm_roi.png}}}
\subcaptionbox{FDK, no prior}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/fbp.png}}
\subcaptionbox{TV, no prior}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/tv_lambdaTV_0.100.png}}
\subcaptionbox{This paper}{\includegraphics[width=0.23\columnwidth]{../images/sprouts/post_TCI/2D/20_views/weighted_prior_kk_30_lambda_prior_2.100000.png}}
\caption{Reconstruction of sprouts from 20 projection views
   (b) has streaky artefacts, (c) has blurred structures and (d) is sharper with significantly less streaky artefacts. 3D reconstructions from 30 cone-beam views can be seen in the supplementary material~\cite{supp_paper}.}
\label{fig:sprouts_2D_results}
\end{figure}
